Beta Diversity through Time and Space

### 2020-09-28 analyses/beta_diversity.Rmd
Updated beta div analysis

@@

# Add site data
beta_mit_data <- beta_site.date %>%
  separate(i, into = c("site1", "date1"), sep = "_") %>%
  separate(j, into = c("site2", "date2"), sep = "_") %>%
  mutate(date1 = ymd(date1),
         date2 = ymd(date2),
         year1 = factor(year(date1)),
         year2 = factor(year(date2)),
         yday1 = yday(date1),
         yday2 = yday(date2)) %>%
  left_join(site_data, by = c("site1" = "site")) %>%
  dplyr::select(site1, year1, yday1, site2, year2, yday2, S, OS, WN, ST, 
                ST.l, ST.h, ST.lh, WN.repl, OS.repl, WN.rich, OS.rich, 
                elev1 = elev.mean, man1 = management, x1 = x, y1 = y) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, year1, yday1, site2, year2, yday2, S, OS, WN, ST, 
                ST.l, ST.h, ST.lh, WN.repl, OS.repl, WN.rich, OS.rich, 
                elev1, elev2 = elev.mean, man1, man2 = management, x1, y1, x2 = x, y2 = y) %>%
  mutate(site1 = factor(site1),
         site2 = factor(site2),
         year1 = factor(year1),
         year2 = factor(year2),
         yday1 = as.numeric(yday1),
         yday2 = as.numeric(yday2),
         elev.diff = abs(elev1 - elev2),
         mean.elev = (elev1 + elev2)/2,
         yday.diff = abs(yday1 - yday2),
         man.diff = case_when(
           man1 == man2 ~ FALSE,
           man1 != man2 ~ TRUE
         )) %>%
  gather(metric, value, -c(site1, year1, yday1, site2, year2, yday2,
                           elev1, elev2, elev.diff, yday.diff, 
                           man1, man2, man.diff, mean.elev, 
                           x1, y1, x2, y2)) %>%
  mutate(metric.class = case_when(
    metric %in% c("WN", "ST", "OS", "S") ~ "aggregate",
    metric %in% c("WN.repl", "WN.rich") ~ "WN.partition",
    metric %in% c("ST.h", "ST.l", "ST.lh") ~ "ST.partition",
    metric %in% c("OS.repl", "OS.rich") ~ "OS.partition"),
    metric = factor(metric, levels = c("WN", "ST", "OS", "ST.h", "ST.l", 
                                       "ST.lh", "WN.repl", "WN.rich", 
                                       "OS.repl", "OS.rich"))
    )
```

### Plot interaction beta-diversity against elevation difference and time difference
```{r}
# Set palette
p <- c("OS" = "darkorange1", "ST" = "green3", 
       "ST.h" = "firebrick1", "ST.l" = "darkorchid", 
       "ST.lh" = "deepskyblue", "WN" = "black")

# Elevation diff
ggplot(filter(beta_mit_data, year1 == year2 &
                site1 != site2 &
                metric != "S" & 
                !metric.class %in% c("OS.partition", "WN.partition")), 
       aes(elev.diff, value, color = metric)) +
  geom_point(alpha = 0.25, pch = 16) +
  geom_smooth() +
  facet_grid(metric.class ~ year1) +
  scale_color_manual(values = p) +
  theme_abyss()

# Time diff
ggplot(filter(beta_mit_data, year1 == year2 &
                site1 != site2 &
                metric != "S" & 
                !metric.class %in% c("OS.partition", "WN.partition")), 
       aes(yday.diff, value, color = metric)) +
  geom_point(alpha = 0.25, pch = 16) +
  geom_smooth() +
  facet_grid(metric.class ~ year1) +
  scale_color_manual(values = p) +
  theme_abyss()

# Elevation diff within 1 wk time diff
p1 <- ggplot(filter(beta_mit_data, year1 == year2 &
                site1 != site2 &
                metric != "S" & 
                !metric.class %in% c("OS.partition", "WN.partition") 
                #& yday.diff <= 7
                ), 
       aes(elev.diff, value, color = metric)) +
  geom_point(alpha = 0.03, pch = 16, size = 0.25) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6)) +
  facet_grid(metric.class ~ year1) +
  scale_color_manual(values = p, name = "partition") +
  xlab("Elevation difference (m)") +
  ylab("β-diversity")
ggsave("./output/beta_elev.png", height = 4.5, width = 6.5, units = "in")

# Time diff within 250 m elev diff
p2 <- ggplot(filter(beta_mit_data, year1 == year2 &
                site1 != site2 &
                metric != "S" & 
                !metric.class %in% c("OS.partition", "WN.partition")
               # & elev.diff <= 250
                ), 
       aes(yday.diff, value, color = metric)) +
  geom_point(alpha = 0.03, pch = 16, size = 0.25) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6)) +
  facet_grid(metric.class ~ year1) +
  scale_color_manual(values = p, name = "partition") +
  xlab("Time difference (days)") +
  ylab("β-diversity") + 
  theme(legend.position="none")
ggsave("./output/beta_time.png", height = 4.5, width = 6.5, units = "in")

png("./output/beta_stack.png", width = 6.5, height = 6.5, units = "in", res = 300)
p1 / p2 + plot_annotation(tag_levels = c("A", "B", "C")) + plot_layout(guides = "collect")
dev.off()

@@

## collate extracted splines  
interaction_splines <- bind_rows(interaction_splines2010, 
                                 interaction_splines2011, 
                                 interaction_splines2012) %>%
  mutate(x.var = case_when(
    x.var == "elev.mean" ~ "Elevation (m)",
    x.var == "Geographic" ~ "Geographic (km)",
    x.var == "yday" ~ "Day of Year"),
    x.var = factor(x.var, levels = c("Geographic (km)", "Elevation (m)", "Day of Year"))
    )

@@

## collate extracted splines  
survey_splines <- bind_rows(survey_splines2010,
                            survey_splines2011, 
                            survey_splines2012) %>%
  mutate(x.var = case_when(
    x.var == "elev.mean" ~ "Elevation (m)",
    x.var == "Geographic" ~ "Geographic (km)",
    x.var == "yday" ~ "Day of Year"),
    x.var = factor(x.var, levels = c("Geographic (km)", "Elevation (m)", "Day of Year"))
    )

@@

### GDM: BB observations; this won't work in it's present state because I cannot 
```{r}
# Create functions for prepping data
biodata_prep_BB <- function(net, y) {
  net %>%
    filter(year == y) %>%
    group_by(site, date, bb.sp) %>%
    summarize(count = n()) %>%
    mutate(count = rep(1, n())) %>% # this makes the data binary, which I think is what I want here; otherwise, just comment out
    spread(bb.sp, count) %>%
    replace(is.na(.), 0) %>%
    mutate(yday = yday(date)) %>%
    unite(site.day, site, yday, sep = "_", remove = "TRUE") %>%
    ungroup() %>%
    dplyr::select(-date)
}

@@

## collate extracted splines  
bb_splines <- bind_rows(bb_splines2010,
                            bb_splines2011, 
                            bb_splines2012) %>%
  mutate(x.var = case_when(
    x.var == "elev.mean" ~ "Elevation (m)",
    x.var == "Geographic" ~ "Geographic (km)",
    x.var == "yday" ~ "Day of Year"),
    x.var = factor(x.var, levels = c("Geographic (km)", "Elevation (m)", "Day of Year"))
    )

@@

>>>>

### Get beta-diversity for BBs and plants (not their interactions)
```{r}
bb_beta <- net %>%
  mutate(yday = yday(date)) %>%
  group_by(site, yday, year, bb.sp) %>%
  summarize(count = n()) %>%
  unite(id, c(site, yday, year), sep = "_") %>%
  pivot_wider(names_from = bb.sp, values_from = count) %>%
  replace(is.na(.), 0) %>%
  column_to_rownames("id") %>%
  vegdist(method = "bray", binary = TRUE) %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column(var = "id1") %>%
  pivot_longer(cols = -id1, names_to = "id2", values_to = "BB") %>%
  separate(id1, into = c("site1", "yday1", "year1"), sep = "_") %>%
  separate(id2, into = c("site2", "yday2", "year2"), sep = "_")

fl_beta <- survey %>%
  filter(flower.cover > 0) %>%
  mutate(yday = yday(date),
         year= year(date)) %>%
  group_by(site, yday, year, plant.sp) %>%
  summarize(count = sum(flower.cover)) %>%
  unite(id, c(site, yday, year), sep = "_") %>%
  pivot_wider(names_from = plant.sp, values_from = count) %>%
  replace(is.na(.), 0) %>%
  column_to_rownames("id") %>%
  vegdist(method = "bray", binary = TRUE, upper = TRUE) %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column(var = "id1") %>%
  pivot_longer(cols = -id1, names_to = "id2", values_to = "FL") %>%
  separate(id1, into = c("site1", "yday1", "year1"), sep = "_") %>%
  separate(id2, into = c("site2", "yday2", "year2"), sep = "_")

species_beta <- full_join(bb_beta, fl_beta) %>%
  left_join(site_data, by = c("site1" = "site")) %>%
  dplyr::select(site1, year1, yday1, site2, year2, yday2, FL, BB,
                elev1 = elev.mean, x1 = x, y1 = y) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, year1, yday1, site2, year2, yday2, FL, BB,
                elev1, elev2 = elev.mean, x1, y1, x2 = x, y2 = y) %>%
  mutate(site1 = factor(site1),
         site2 = factor(site2),
         year1 = factor(year1),
         year2 = factor(year2),
         yday1 = as.numeric(yday1),
         yday2 = as.numeric(yday2),
         elev.diff = elev1 - elev2,
         mean.elev = (elev1 + elev2)/2,
         yday.diff = yday1 - yday2,
         yday.diff.abs = abs(yday1 - yday2)) %>%
  pivot_longer(cols = c(BB, FL), names_to = "group", values_to = "beta") %>%
  filter(year1 == year2 &
           elev.diff >= 0) # this simply deduplicates the data so that site1 vs site2 and site2 vs. site 1 are not both included; only works because elev.diff is unique for each site pair
```

@@

# Add site data
beta_mit_data <- beta_site.date %>%
  separate(i, into = c("site1", "date1"), sep = "_") %>%
  separate(j, into = c("site2", "date2"), sep = "_") %>%
  mutate(date1 = ymd(date1),
         date2 = ymd(date2),
         year1 = factor(year(date1)),
         year2 = factor(year(date2)),
         yday1 = yday(date1),
         yday2 = yday(date2)) %>%
  left_join(site_data, by = c("site1" = "site")) %>%
  dplyr::select(site1, year1, yday1, site2, year2, yday2, S, OS, WN, ST, 
                ST.l, ST.h, ST.lh, WN.repl, OS.repl, WN.rich, OS.rich, 
                elev1 = elev.mean, man1 = management, x1 = x, y1 = y) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, year1, yday1, site2, year2, yday2, S, OS, WN, ST, 
                ST.l, ST.h, ST.lh, WN.repl, OS.repl, WN.rich, OS.rich, 
                elev1, elev2 = elev.mean, man1, man2 = management, x1, y1, x2 = x, y2 = y) %>%
  mutate(site1 = factor(site1),
         site2 = factor(site2),
         year1 = factor(year1),
         year2 = factor(year2),
         yday1 = as.numeric(yday1),
         yday2 = as.numeric(yday2),
         elev.diff = abs(elev1 - elev2),
         mean.elev = (elev1 + elev2)/2,
         yday.diff.abs = abs(yday1 - yday2),
         man.diff = case_when(
           man1 == man2 ~ FALSE,
           man1 != man2 ~ TRUE
         )) %>%
  gather(metric, value, -c(site1, year1, yday1, site2, year2, yday2,
                           elev1, elev2, elev.diff, yday.diff.abs, 
                           man1, man2, man.diff, mean.elev, 
                           x1, y1, x2, y2)) %>%
  mutate(metric.class = case_when(
    metric %in% c("WN", "ST", "OS", "S") ~ "aggregate",
    metric %in% c("WN.repl", "WN.rich") ~ "WN.partition",
    metric %in% c("ST.h", "ST.l", "ST.lh") ~ "ST.partition",
    metric %in% c("OS.repl", "OS.rich") ~ "OS.partition"),
    metric = factor(metric, levels = c("WN", "ST", "OS", "ST.h", "ST.l", 
                                       "ST.lh", "WN.repl", "WN.rich", 
                                       "OS.repl", "OS.rich"))
    )
```

### Plot interaction beta-diversity against elevation difference and time difference
```{r}
# Set palette
p <- c("OS" = "deepskyblue", "ST" = "green3", 
       "ST.h" = "firebrick1", "ST.l" = "darkorchid", 
       "ST.lh" = "darkorange1", "WN" = "black")

# BB + FL elevation diff
g1 <- ggplot(filter(species_beta, site1 != site2 & yday.diff.abs <= 5), aes(elev.diff, beta)) +
  geom_point(alpha = 0.05, pch = 16) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6)) +
  facet_grid(group ~ year1) +
  xlab("Elevation difference (m)") +
  ylab("β-diversity")

# BB + FL time diff
g2 <- ggplot(filter(species_beta, site1 == site2 & yday.diff > 0), aes(yday.diff.abs, beta)) +
  geom_point(alpha = 0.05, pch = 16) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6)) +
  facet_grid(group ~ year1) +
  xlab("Time difference (days)") +
  ylab("β-diversity")

# BB * FL elevation diff
g3 <- ggplot(filter(beta_mit_data, year1 == year2 &
                      site1 != site2 &
                      metric != "S" & 
                      !metric.class %in% c("OS.partition", "WN.partition") & 
                      yday.diff.abs <= 5), 
             aes(elev.diff, value, color = metric)) +
  #geom_point(alpha = 0.03, pch = 16, size = 0.25) +
  geom_point(alpha = 0.05, pch = 16, size = 1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6)) +
  #facet_grid(metric.class ~ year1) +
  facet_wrap(~year1) +
  scale_color_manual(values = p, name = "partition") +
  xlab("Elevation difference (m)") +
  ylab("β-diversity")

# BB * FL time diff
g4 <- ggplot(filter(beta_mit_data, year1 == year2 &
                      site1 == site2 &
                      metric != "S" & 
                      !metric.class %in% c("OS.partition", "WN.partition")), 
             aes(yday.diff.abs, value, color = metric)) +
  #geom_point(alpha = 0.03, pch = 16, size = 0.25) +
  geom_point(alpha = 0.05, pch = 16, size = 1) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 6)) +
  #facet_grid(metric.class ~ year1) +
  facet_wrap(~year1) +
  scale_color_manual(values = p, name = "partition") +
  xlab("Time difference (days)") +
  ylab("β-diversity") + 
  theme(legend.position="none")

png("./output/beta_stack.png", width = 9, height = 6.5, units = "in", res = 300)
(g1 + g3) / (g2 + g4) + 
  plot_annotation(tag_levels = c("A", "B", "C", "D")) + 
  plot_layout(guides = "collect") & theme_light()
dev.off()

@@

## collate extracted splines  
interaction_splines <- bind_rows(interaction_splines2010, 
                                 interaction_splines2011, 
                                 interaction_splines2012) %>%
  mutate(x.var = case_when(
    x.var == "elev.mean" ~ "Elevation (m)",
    x.var == "Geographic" ~ "Geographic (km)",
    x.var == "yday" ~ "Day of Year"),
    x.var = factor(x.var, levels = c("Geographic (km)", "Elevation (m)", "Day of Year")),
    subject = rep("BB*FL", n())
    )

@@

## collate extracted splines  
survey_splines <- bind_rows(survey_splines2010,
                            survey_splines2011, 
                            survey_splines2012) %>%
  mutate(x.var = case_when(
    x.var == "elev.mean" ~ "Elevation (m)",
    x.var == "Geographic" ~ "Geographic (km)",
    x.var == "yday" ~ "Day of Year"),
    x.var = factor(x.var, levels = c("Geographic (km)", "Elevation (m)", "Day of Year")),
    subject = rep("FL", n())
    )

@@

### GDM: BB observations
```{r}
# Create functions for prepping data
biodata_prep_BB <- function(net, y) {
  net %>%
    filter(year == y) %>%
    group_by(site, date, bb.sp) %>%
    summarize(count = n()) %>%
    mutate(count = rep(1, n())) %>% # this makes the data binary, which I think is what I want here; otherwise, just comment out
    spread(bb.sp, count) %>%
    replace(is.na(.), 0) %>%
    mutate(yday = yday(date)) %>%
    unite(site.day, site, yday, sep = "_", remove = "TRUE") %>%
    ungroup() %>%
    dplyr::select(-date)
}

@@

## collate extracted splines  
bb_splines <- bind_rows(bb_splines2010,
                            bb_splines2011, 
                            bb_splines2012) %>%
  mutate(x.var = case_when(
    x.var == "elev.mean" ~ "Elevation (m)",
    x.var == "Geographic" ~ "Geographic (km)",
    x.var == "yday" ~ "Day of Year"),
    x.var = factor(x.var, levels = c("Geographic (km)", "Elevation (m)", "Day of Year")),
    subject = rep("BB", n())
    )

@@

### GDM all together now
```{r}
splines <- bind_rows(interaction_splines, bb_splines, survey_splines) %>%
  mutate(subject = factor(subject))

ggplot(splines, aes(predictor, response, color = subject)) +
  geom_line() +
  facet_grid(year ~ x.var, scales = "free_x") +
  xlab(NULL) +
  ylab("f(x)") +
  theme_light(12)
ggsave("./output/GDM.png")
```
